---
title: "Lab 1"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Variance of normal RV-s
Let's compute the sample variance of normal random variables by Monte Carlo (MC) simulations.
With sample size *n = 1000*, let's generate n samples to work with.

```{r sample}
n <- 1000
sample <- rnorm(n)
sample_mean <- mean(sample)
```

The mean of this sample is `r sample_mean`, which is unsurprisingly very close to 0. Next, let's explore different ways of finding variance of this sample. 

```{r variance}
sample_variance <- mean((sample - sample_mean)^2)
shortcut_variance <- mean((sample^2)) - mean(sample)^2
var_variance <- var(sample)
```

Following the formula **Var X = E[(X - E X)^2]**, we get **`r sample_variance`**. Using the shortcut formula **Var X = E x^2 - (E X)^2**, we get the same value! Computing variance using the canned `var` function gives us **`r var_variance`**, which is slightly larger than the other variances we computed.

Now let's dive more into our sample, finding the standard deviation and confidence interval bounds.

```{r sd}
# Finding standard deviation
sample_sd <- sd(sample)

# Finding confidence interval
lower <- sample_mean - 1.96 * sample_sd
upper <- sample_mean + 1.96 * sample_sd
conf_interval <- ((length(sample[sample > upper]) + length(sample[sample < lower])) / n)
```


Using the `sd` function, we get `r sample_sd` as the standard deviation of the sample. To find the percentage of our sample that falls outside of the range of the 95% confidence interval, or **[X - 1.96 * sd X, X + 1.96 * sd X]**, we first find the lower and upper bounds of that interval. This happened to have a lower bound of `r lower` and upper bound of `r upper`. The percentage of our values that fall outside of these bounds is about **`r conf_interval * 100`%**, which is unsurprisingly close to 5%.

We can plot a histogram our our sample, showing the mean and confidence interval. Here we see a binomial distribution of the sample numbers, with a minimal amount (5%) of the data falling outside of the 95% confidence interval. Most values are centered around the mean, while less trickle off and away. The mean of the distribution is nearly zero.

```{r histogram, echo=FALSE}
hist(sample, col = "pink")
abline(v = sample_mean, col = "blue")
abline(v = lower, col = "red")
abline(v = upper, col = "red")
```

## Variance of means

Now we move a step further and compute the variance of means: instead of picking n random numbers, now we pick n random numbers m time, and each calculate their mean. Afterwards, we'll see how the variance of means will change when we change n. For this example, we will have *n = 3* and *m = 1000*.

```{r means}
n <- 3
m <- 1000

# Find means of m samples of n
means <- sapply(replicate(m, sapply(n, rnorm, simplify = FALSE)), mean)
variance <- var(means) # variance of means

# What happens if we make sample size larger?
new_n <- n * 100
new_means <- sapply(replicate(m, sapply(new_n, rnorm, simplify = FALSE)), mean)
new_variance <- var(new_means)
# The variance is much smaller!
```

The variance of our sample of means is **`r variance`**. When sample size increases, variance decreases. For example, when we change our sample size n to be 10x larger (n=300), our new variance is **`r new_variance`**!

```{r meanssd, echo=FALSE}
# How much smaller will standard deviation be if the sample is 10x larger?
first_sd <- sd(means)

second_n <- n * 10
second_means <- sapply(replicate(m, sapply(second_n, rnorm, simplify = FALSE)), mean)
second_sd <- sd(second_means)

difference <- round(first_sd / second_sd, 2)
# The new standard deviation will be around 3x smaller than the first standard deviation
```

*How much smaller will the standard deviation be if the sample size is 10x larger?* The first standard deviation of this sample comes out to be `r first_sd`, and the new standard deviation when changing the sample size to be 10x larger is `r second_sd`. The new standard deviation appears to consistently be around 3x smaller than the first one. In this example, the new standard deviation is **`r difference`x** smaller!
