---
title: "Info 371 Lab 3"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Linear Regression

This lab is about linear regression, dummy variables, and interaction eects. In many ways this is about feature engineering and related interpretation.

This lab is based on World Value Survey data (see http://www.worldvaluesurvey.org/wvs.jsp), in particular on a small subset of the 6th wave of the survey for the US (collected in 2011) and China (2013).

Our task in this lab is to estimate which factors influence choice: how much freedom people think they have in their life, and more importantly, to interpret the results.


### Gender and Country

First, let's take very simple look and estimate a linear regression model where freedom of choice depends on the country. Note that a larger number of freedom means the respondents feel they are more free in their decisions.

``` {r echo=FALSE}
wvs <- read.delim("wvs.csv.bz2")

m <- lm(choice ~ country, data=wvs)
summary(m)
```

Chinese citizens reported: 7.15841
American citizens reported: 7.76369

Linear regression formula: 
**Choice = 7.15841 + 0.60528.US**

American respondents feel more free in their choices than Chinese respondents by **0.60528**. The measurement unit is numeric scale between 1 and 10 with 1 = "no choice at all", 10 = "a great deal of choice."

Multiple R-squared: 0.02495 

R^2 tells us how much variation in DV can be explained by the different levels of the IV (e.g., China and US). We can conclude that *country* does not correlate much with *choice*.



### Sex and Country

``` {r echo=FALSE}
m2 <- lm(choice ~ sex + country, data=wvs)
summary(m2)
```


Linear regression formula: 

**Choice = 7.18674 + (-0.05748.Male) + 0.60473.US**


Based on our estimations, lets find the level of choice freedom for...

1. American men: `r 7.18674 + 0.60473 - 0.05748`
2. American women: `r 7.18674 + 0.60473`
3. Chinese men: `r 7.18674 - 0.05748`
4. Chinese women: `r 7.18674`


### Now add interaction: sex x country

Now let's add the interaction effect, sex x country. If sex and country are numbers, it amounts to just adding a product of these two. In our case these are categorical variables, automatically transformed to dummies by R (and by statsmodels.formula.api in python, and many other software too). The standard R-style formula uses asterisk, the multiplication operator, *, to denote interaction.

``` {r echo=FALSE}
m3 <- lm(choice ~ sex * country, data = wvs)
summary(m3)
```



Linear regression formula:

**Choice = 7.058 + 0.85302.US + 0.20366.Male -0.50857.US.Male**

Based on our estimations, lets find the level of choice freedom for...

1. American men: `r 7.058 + 0.85302 + 0.20366 - 0.50857`
2. American women: `r 7.05803 + 0.85302`
3. Chinese men: `r 7.05803 + 0.20466`
4. Chinese women: `r 7.05803`

Looking at our estimations, we can observe that *in the US, women have a slightly higher level of choice freedom than men, while in China, women have a slightly lower level of choice freedom.*


